---
title: "Statistical assignment 3"
author: "Helen Clara Retzlaff, Candidate number: 142918"
date: "12.02.2020"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
# Please note these options.
# This tells R Markdown that we want to show code in the output document.
knitr::opts_chunk$set(echo = TRUE)
# Switching off messages in the output document. 
knitr::opts_chunk$set(message = FALSE)
# Switching on caching to make things faster (don't commit cache files on Github).
knitr::opts_chunk$set(cache = TRUE)
```

In this assignment we will explore political interest (*vote6*) and how it changes over time.

## Read data

First we want to read and join the data for the first 7 waves of the Understanding Society. (Wave 8 does not have a variable for political interest). We only want five variables: personal identifier, sample origin, sex, age and political interest. It is tedious to join all the seven waves manually, and it makes sense to use a loop in this case. Since you don't yet know about iteration I'll provide the code for you; please see the explanation of the code here: http://abessudnov.net/dataanalysis3/iteration.html.

The only thing you need to do for this code to work on your computer is to provide a path to the directory where the data are stored on your computer.

```{r}
library(tidyverse)
#install.packages("data.table") as I already had it installed
library(data.table)

# data.table is faster compared to readr so we'll use it in this case (the function fread()). You need to install this package first to be able to run this code.

# create a vector with the file names and paths

files <- dir(
             # Select the folder where the files are stored.
             "/Users/helenretzlaff/Documents/Data 3/datan3_2020/data/UKDA-6614-tab/tab",
             # Tell R which pattern you want present in the files it will display.
             pattern = "indresp",
             # We want this process to repeat through the entire folder.
             recursive = TRUE,
             # And finally want R to show us the entire file path, rather than just
             # the names of the individual files.
             full.names = TRUE)

# Select only files from the UKHLS.
files <- files[stringr::str_detect(files, "ukhls")]
files

# create a vector of variable names
vars <- c("memorig", "sex_dv", "age_dv", "vote6")

for (i in 1:7) {
        # Create a vector of the variables with the correct prefix.
        varsToSelect <- paste(letters[i], vars, sep = "_")
        # Add pidp to this vector (no prefix for pidp)
        varsToSelect <- c("pidp", varsToSelect)
        # Now read the data. 
        data <- fread(files[i], select = varsToSelect)
        if (i == 1) {
                all7 <- data  
        }
        else {
                all7 <- full_join(all7, data, by = "pidp")
        }
        # Now we can remove data to free up the memory.
        rm(data)
} 

#the data has 83462 observations and 29 variables 

```

## Reshape data (20 points)

Now we have got the data from all 7 waves in the same data frame **all7** in the wide format. Note that the panel is unbalanced, i.e. we included all people who participated in at least one wave of the survey. Reshape the data to the long format. The resulting data frame should have six columns for six variables.

```{r}
Long <- all7 %>%
  pivot_longer(a_memorig:g_vote6,names_to="variable",values_to = "value") %>% 
  separate(variable,into = c("wave","variable"),sep = "_",extra = "merge") %>% 
  pivot_wider(names_from = variable,values_from = value)
  
Long
```

## Filter and recode (20 points)

Now we want to filter the data keeping only respondents from the original UKHLS sample for Great Britain (memorig == 1). We also want to clean the variables for sex (recoding it to "male" or "female") and political interest (keeping the values from 1 to 4 and coding all negative values as missing). Tabulate *sex* and *vote6* to make sure your recodings were correct. 

```{r}

Long <- Long %>%
        filter(memorig==1) %>%
        mutate(sex_dv = ifelse(sex_dv == 2, "female",
                           ifelse(sex_dv == 1, "male", NA))) %>%
        mutate(vote6 = ifelse(vote6<0,NA,vote6))

Long %>%
        count(sex_dv)

```

```{r}

Long %>%
        count(vote6)

```

## Calculate mean political interest by sex and wave (10 points)

Political interest is an ordinal variable, but we will treat it as interval and calculate mean political interest for men and women in each wave.

```{r}
meanVote6 <- Long %>%
        filter(!is.na(sex_dv)) %>%
        group_by(sex_dv,wave) %>%
        summarise(
                meanVote6 = mean(vote6, na.rm = TRUE)
        )
        
meanVote6
```

## Reshape the data frame with summary statistics (20 points)

Your resulting data frame with the means is in the long format. Reshape it to the wide format. It should look like this:

| sex_dv  | a  | b  | c  | d  |  e | f  | g  |
|---      |--- |--- |--- |--- |--- |--- |--- |
| female  |    |    |    |    |    |    |    |
| male    |    |    |    |    |    |    |    |

In the cells of this table you should have mean political interest by sex and wave.

Write a short interpretation of your findings.

```{r}
meanVote6 %>%
        pivot_wider(names_from = wave,values_from = meanVote6)
```

Interpretation: 

These findings suggest that political interest remained relatively constant throughout all 7 waves, for both women and men. For the former, political interest varied betwen 2.72 in wave 7 (highest) and 2.88 in wave 4 (lowest). For the latter, interest varied between 2.41 in wave 7 (highest) and 2.55 in wave 4 (lowest). 

Hence, men in this survey are slightly more interested in politics as 1 indicates being very interested in politics and 4 being not at all interested. Both sexes lie in the "mid-range" between beeing fairly and not very interested in politics. 
There is some variation for both sexes and their highest and lowest interest in politics are in the same waves. This could be due to external factors like frustration with politics or a dramatic event. Wave 7 was conducted between 2015-16 where the Brexit debate and referendum occured, so this could be an explanation for the increase in political interest for both sexes. 


## Estimate stability of political interest (30 points)

Political scientists have been arguing how stable the level of political interest is over the life course. Imagine someone who is not interested in politics at all so that their value of *vote6* is always 4. Their level of political interest is very stable over time, as stable as the level of political interest of someone who is always very interested in politics (*vote6* = 1). On the other hand, imagine someone who changes their value of *votes6* from 1 to 4 and back every other wave. Their level of political interest is very unstable.

Let us introduce a measure of stability of political interest that is going to be equal to the sum of the absolute values of changes in political interest from wave to wave. Let us call this measure Delta. It is difficult for me to typeset a mathematical formula in Markdown, but I'll explain this informally.

Imagine a person with the level of political interest that is constant over time: {1, 1, 1, 1, 1, 1, 1}. For this person, Delta is zero.

Now imagine a person who changes once from "very interested in politics" to "fairly interested in politics": {1, 1, 1, 1, 2, 2, 2}. For them, Delta = (1 - 1) + (1 - 1) + (1 - 1) + (2 - 1) + (2 - 2) + (2 - 2) = 1.

Now imagine someone who changes from "very interested in politics" to "not at all interested" every other wave: {1, 4, 1, 4, 1, 4, 1}. Delta = (4 - 1) + abs(1 - 4) + (4 - 1) + abs(1 - 4) + (4 - 1) + abs(1 - 4) = 3 * 6 = 18.

Large Delta indicates unstable political interest. Delta = 0 indicates a constant level of political interest.

Write the R code that does the following.

1. To simplify interpretation, keep only the respondents with non-missing values for political interest in all seven waves.
2. Calculate Delta for each person in the data set.
3. Calculate mean Delta for men and women.
4. Calculate mean Delta by age (at wave 1) and plot the local polynomial curve showing the association between age at wave 1 and mean Delta. You can use either **ggplot2** or the *scatter.smooth()* function from base R.
5. Write a short interpretation of your findings.

```{r}

#1. get rid of NA´s in variable for political interest

balanced<- all7%>% 
  filter(complete.cases(a_vote6,b_vote6,c_vote6,d_vote6,e_vote6,f_vote6,g_vote6))

#the balanced panel has 18731 observations and 29 variables 

#2. Calculate Delta for each person in the data set

deltaEach<-balanced %>% 
  mutate(deltaEach=abs(b_vote6-a_vote6)+abs(c_vote6-b_vote6)+abs(d_vote6-c_vote6)+abs(e_vote6-d_vote6)+abs(f_vote6-e_vote6)+abs(g_vote6-f_vote6))

deltaEach %>% 
  count(deltaEach)

# most individuals in the survey have a delta between 0 and 5, indicating some instability of political interest over the waves but not a lot. This indicates that it is unlikely for most people to become very interested in politics if they were not interested at all before. 
```


```{r}

#3. Calculate mean Delta for men and women

delta.gender <- deltaEach%>%
  group_by(a_sex_dv) %>%
  summarise(
    DeltaGender = mean(deltaEach, na.rm = TRUE)
  )

#I used a_sex_dv as the sex variable because gender is a stable variable and does not vary between the waves. 

delta.gender

#the table indicates a slightly higher instability of political interest for men (1) than for women (2). 

```


```{r}

#4. Calculate mean Delta by age (at wave 1) and plot the local polynomial curve showing the association between age at wave 1 and mean Delta. You can use either **ggplot2** or the *scatter.smooth()* function from base R.

delta.age <- deltaEach%>%
  group_by(a_age_dv) %>%
  summarise(
    DeltaAge = mean(deltaEach, na.rm = TRUE)
  )

delta.age

```


```{r}
  
#4.1 Plot

delta.age %>%  
    ggplot(aes(x=a_age_dv, y=DeltaAge, group=1)) + 
    geom_point(color='red',   alpha=0.3,size=3)+
    stat_smooth(method='lm', formula = y~poly(x,2))+xlab("age at wave 1")+ylab("mean Delta")
  
```


#5. Interpretation 

The higher Delta, the more unstable political interest. The graph indicates that for younger people (up until the age of about 35), political interest is less stable than for people between the age of 35 and 60. in that range, political interest seems stable as Delta is not high or fluctuates. As people get old, Delta increases again, indicating unstable political interest. However, this result might be skewed due to one extreme outlier at the upper end of the spectrum. 

